---
title: "Newton Raphson in a Poisson model"
author: "YOUR NAME"
date: "5/12/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(fastR2)
library(tidyverse)
```

1. An instructor believes that the number of students who arrive late to class should follow a Poisson distribution. The table below indicates the number of such students in 10 consecutive classes.

\begin{table}[h]
\centering
\begin{tabular}{cccccc}\hline
number of late students & 0 & 1 & 2 & 3 & 4 \\
frequency & 3& 3 & 2 & 1 & 1 \\ \hline
\end{tabular}
\end{table}

Let $X_1, X_2, \dots, X_{10}$ denote the number of students who were late to each of the 10 classes. A reasonable model is 
$$X_{1}, X_2, \dots, X_{10} \overset{i.i.d.}{\sim} Pois(\lambda).$$

a. Write the log-likelihood function for $\lambda$ based on this model. What is the ML estimator of $\lambda$? (Write it out on paper and then type it using Latex for practice.)

b. Make a graph of the log likelihood function for this data set. (Take a look at practice problems 1.1 from onenote for code Harshil gave you in section for hints on how to approach this.)

```{r}



```

c. Using R as a calculator, work out three iterations of the Newton-Raphson algorithm for finding the MLE $\widehat{\lambda}$ using a starting guess of $\lambda_0=1$.

d. Use `maxLik` and calculate the MLE of $\lambda$ using Newton Raphson. (You will need to write a function, `loglik.pois` to calculate the log likelihood from a. )