---
title: "Robustness of t confidence intervals"
author: "YOUR NAME"
date: "4/13/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(mosaic)
library(infer)
library(fastR2)

```

In this investigation, you will examine the coverage probability of t confidence intervals when the data $X_1, X_2, \dots, X_n$ do not arise from a normal distribution.  

1. 
a. Use `CISim` from the **mosaic** package to calculate the coverage probability of 95% confidence intervals for 1,000 samples of size $n=5, 10,  30$ from an exponential distribution with rate $\lambda = 1/10$. Be sure to set the seed to 16313. 

```{r simulate-exponential}
#code here

```

b. Is the coverage close to 95%? If not, do your intervals miss the true mean because they tend to be below it? Or above it? 

2. In the following code chunk, I have  written a function called `tcalc` which calculates the t-ratios 
$$t = \frac{ \sqrt{n}( \bar{X} - 10)}{S}$$
a sample of size $n$. Using the `replicate` function, I then repeatedly evaluate `tcalc` 1,000 times and save the result in a tibble called `simdata`.

  Use `ggplot` to make a histogram of the t-ratios. Does the distribution look like a t? If not, what's the problem with it. 

```{r calculate-tratio}

tcalc <- function(x, mu0){
  #calculate t = sqrt(n) (xbar - mu0)/s 
  n = length(x)
  xbar = mean(x)
  s = sd(x)
  t = sqrt(n)*(xbar - mu0)/s
  return(t)
}

simdata <- tibble(
  #generate X1, X2, ..., X5 from Exp(lambda = 1/10) and calculate tcalc
  #replicate 1000 times
  t = replicate( n = 1000, 
                 expr = tcalc(rexp(n = 5, rate=1/10), mu0 = 10)  )
  
)


```

3. Use your histogram from 2. to better understand the behavior of the confidence intervals from 1.



4. Repeat exercises 1 and 2 for data which arises from a Uniform(-1,1). 