---
title: "Quiz 04/11 R Code"
output:
  pdf_document: 
    fig_width: 5
    fig_height: 3
date: '2022-04-11'
---

```{r,message=FALSE,warning=FALSE,echo=F}
library(tidyverse)
library(ggridges)
library(infer)
library(fastR2)
```

# Problem 1.2

Construct 95% CI's for the mean ratio of Sepal.Length and Sepal.Width for each species found in the Iris dataset.

First we may do some exploratory data analysis on these values to determine if a t-test is appropriate. 

Using glimpse is always a good way to examine what your data looks like before moving on to visualizations.

```{r}
glimpse(iris)
```

For our visualizations may begin with a faceted histogram. (For binwidth see: Sturges rule or alternatively use (range of data)/sqrt(n) ).

```{r}
#pipe data into ggplot
iris %>%
  #mapping defines what variables
  #y=..density.. is optional, it just normalizes the histogram heights for easier comparisons
  ggplot(mapping = aes(x=Sepal.Length/Sepal.Width, y=..density..))+
  #geom_histogram layer adds histogram with specified binwidth
  geom_histogram(binwidth = .1)+
  #facet_wrap by categorical variable for divided plots by category
  facet_wrap(~Species) +
  ggtitle('Faceted Histogram of Sepal Length to Width Ratio by Species')
```

From these plots, we see some symmetry, however it is difficult to truly see the shape of distributions. We can also crate a box-plot to view if there are a large amount of outliers. 

```{r}
#fill creates coloring + legend
ggplot(iris,aes(x=Sepal.Length/Sepal.Width,y=Species,fill=Species))+
  #boxplot layer adds boxplot
  geom_boxplot()+
  ggtitle("Box-plot of Sepal Length to Width by Species")
```

There are a few outliers, but nothing too drastic so we can choose to keep them in the dataset. If there were a large amount of outliers, some data cleaning may be necessary. 

Let's use a qqplot to now view the normality of our data. 

```{r}
iris %>%  
  #gf_qq graphs qq plots for ~variable and facets by Species using the |
  gf_qq(~Sepal.Length/Sepal.Width | Species) %>% 
  #qqline adds normal qq lines to see how much data differs from a normal dist
  gf_qqline()
```

These all look relatively decent (the points fit the line), but we can view the points visually as well in density ridge plots.

From the ridge plots we can see that the data looks relatively normal and a t-test is appropriate. (Optional Exercise: try to match up what you see in the qqplot with the density ridge plot).

```{r}
ggplot(iris,aes(x=Sepal.Length/Sepal.Width,y=Species,fill=Species))+
  geom_density_ridges()+
  ggtitle("Density Ridge plot of Sepal Length to Width by Species")
```

We can now calculate some summary statistics to find these confidence intervals. 

```{r}
species_stats <-  iris %>% 
  #group_by categorical variable (Species so we can get info for 3 different CI's)
  group_by(Species) %>% 
  #use summarize to find summary stats
  summarize(ratio_mean = mean(Sepal.Length/Sepal.Width),
            ratio_sd = sd(Sepal.Length/Sepal.Width),
            #n() finds sample size
            n())
species_stats
```

With this we have all the information we need to construct a 95% confidence interval using the t-distribution.

$$
\bar{X} \pm t_{\alpha/2=.05/2,n-1=49}\frac{S}{\sqrt{n}}
$$
We obtain the following 95% confidence intervals:

```{r}
n <-  50
alpha <-  .05
t_95 <-  qt(1-alpha/2,df=n-1)

#storing S and X_bar values in vectors
sd_species <- species_stats$ratio_sd
mean_species <- species_stats$ratio_mean

#calculate bounds according to formula
lower_CI <- mean_species - t_95*(sd_species/sqrt(n))
upper_CI <- mean_species + t_95*(sd_species/sqrt(n))
data.frame(species_stats$Species,lower_CI,upper_CI)
```

To confirm our answers, we may use t_test from the `infer` package to calculate the intervals as well. 

```{r}
#For Setosa

iris %>% filter(Species=='setosa') %>% 
  #mutate new variable (create new column)
  mutate(ratio = Sepal.Length/Sepal.Width) %>%
  #within t_test we set a response=variable for CI, and a conf.level
  #response=Sepal.Length/Sepal.Width would not work, as the variable must be in the df for t_test
  #so we require the mutate
  infer::t_test(response=ratio, conf.level=.95) %>% 
  #selecting/mutating variables for cleaner output
  mutate(Species='setosa CI') %>% 
  select(Species, lower_ci,upper_ci)

#For Versicolor

iris %>% filter(Species=='versicolor') %>% 
  mutate(ratio = Sepal.Length/Sepal.Width) %>% 
  infer::t_test(response=ratio, conf.level=.95) %>% 
  mutate(Species='versicolor CI') %>% 
  select(Species, lower_ci,upper_ci)

#For Virginica

iris %>% 
  filter(Species=='virginica') %>% 
  mutate(ratio = Sepal.Length/Sepal.Width) %>% 
  infer::t_test(response=ratio, conf.level=.95) %>% 
  mutate(Species='virginica CI') %>% 
  select(Species, lower_ci,upper_ci)

```



```

