---
title: "Quiz0516"
output: pdf_document
date: '2022-05-16'
---

```{r,messages=F,warnings=F,include=F}
library(MASS)
library(maxLik)
library(ggplot2)
```

## Practice Problem 3.2.2

First we may graph the density of these duration times using ggplot. For the binwidth we can use the rule of thumb, range(x)/sqrt(n). From our plots we clearly observe bimodal behavior, so a mixture model does make sense. 


```{r}
rule_of_thumb = (max(geyser$duration)-min(geyser$duration))/sqrt(nrow(geyser))

ggplot(geyser,aes(x=duration))+
  geom_histogram(binwidth=rule_of_thumb)
```

First we can create our likelihood function, denoted lik.norm. 

```{r}
#define liklihood with vector of parameters arg, and data (x)
lik.norm <- function(parameters, x){
  #extract all the relevant parameters
  alpha = parameters[1]
  mu1 = parameters[2]
  mu2 = parameters[3]
  sd1 = parameters[4]
  sd2 = parameters[5]
  #set bounds for the ranges of these variables
  if(alpha < 0 | alpha > 1 | sd1 < 0 | sd2 < 0){
    NA
  }
  #output the density
  else{
     (alpha*dnorm(x,mu1,sd1))+(1-alpha)*dnorm(x,mu2,sd2)
  }
}
```

Next, we can denote our log-likelihood function, by taking the sum of logs of the likelihood above. 

```{r}
#create log-likelihood by summing the logs of the previous func.
loglik.norm <- function(parameters,x)
  {
  sum(log(lik.norm(parameters,x)))
  }
```

Next we may use the maxLik function to maximize this log-likeloihood using the Newton-Raphson algorithm. For our initalizations, we can set mu1 = mean(duration)-1, mu2 = mean(duration)+1, sd1=sd2=sd(duration). For alpha we will can use equal mixture for our starting point, so alpha=0.5.

```{r}
#initialize
m <- mean(geyser$duration)
s <- sd(geyser$duration)
alpha_init <- 0.5

#save the estimates so we can extract them later. 
estimates <- maxLik(logLik = loglik.norm, 
       start = c(alpha_init,m-1,m+1,s,s),
       method = "NR",
       tol = 1e-4,
       x = geyser$duration
       )
estimates
```

We can now overlay our MLE estimate over the sample distribution. 

Using geom_line:

```{r}
ggplot(geyser,aes(x=duration))+
  geom_histogram(binwidth=rule_of_thumb, aes(y=..density..))+
  #overlay density using the lik.norm function and geom_line
  geom_line(aes(y=lik.norm(estimates$estimate,duration)),col='red')+
  labs(title='Geyser Duration times with Normal Mixture MLE overlayed',
  x='Duration',
  y='Density')
```
Using geom_function:

```{r}
ggplot(geyser,aes(x=duration))+
  geom_histogram(binwidth=rule_of_thumb, aes(y=..density..))+
  #pass in function, and args as a list.
  geom_function(fun=lik.norm, args = list(parameters = estimates$estimate),col='red')+
  labs(title='Geyser Duration times with Normal Mixture MLE overlayed',
  x='Duration',
  y='Density')
```



















